---
title: "ggro-poster-dataviz"
format: html
editor: visual
author: "Jordan Sibley"
date: last-modified
editor_options: 
  chunk_output_type: console
---

## Set Up

```{r}
#| output: false
#| quiet: true
#| code-fold: true
#| code-summary: "Show the code"
# Libraries 
library(tidyverse)
library(readxl)
library(showtext)
library(here)
library(janitor)
library(lubridate)
library(showtext)
library(statebins) # for heatmap
library(treemapify) # for treemap
library(ggridges) # ridgeline density 
library(waffle) # waffle plot 
library(scales) # commas 
library(ggtext)
library(svglite) # for exporting svg

# Font 
font_add_google(name = "Lato", family = "Lato")
showtext_auto(enable = TRUE)

```

```{r}
#| output: false
#| quiet: true
#| code-fold: true
#| code-summary: "Show the code"
# Data 

# ---- Hawkwatch data sets ----
# Total dataset
hw_full_raw <- read_csv(here("data", "hawk-watch", "Hawkwatch_Final.csv"))

# ---- Banding data ----
banding_raw <- read_excel(here("data", "banding", "BandingFlat01.14.26.xlsx"), sheet = 1)

```

## Exploration and Cleaning

### Hawk Watch Data

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# View dataset
glimpse(hw_full_raw)

# Write function to print out unique values and summary stats 
qc_summary <- function(data) {
  for (col in names(data)) {
    cat("\n---", col_name, "---\n")
    
    if (is.character(data[[col_name]])) {
      unique_vals <- unique(data[[col_name]])
      cat("Unique values:\n")
      print(unique_vals)
    } else if (is.numeric(data[[col_name]])) {
      cat("Summary:\n")
      print(summary(data[[col_name]]))
    } else {
      cat("Column is of type:", class(data[[col_name]]), "- not character or numeric.\n")
    }
  }
}

# Check quality of hawkwatch full 
#qc_summary(hw_full_raw)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Clean hawk watch data 
hw_full_clean <- hw_full_raw |> 
  # clean column names 
  clean_names(case = "upper_camel") |> 
  # remove columns 
  select(c(-TeamName, -Dayleader, -NumberObservers, -NumberVisitors, -WindDirection, -WindSpeed, -ProofLevelName, -IncludeInTotals, -YearDay, -MonthDate, -LocalStartTime, - LocalEndTime, -UtcStartTime, -UtcEndTime)) |> 
  # filter to birds of interest 
  filter(CommonName %in% c("American Kestrel", "Bald Eagle", "Broad-winged Hawk", "Cooper's Hawk",	"Ferruginous Hawk", "Golden Eagle", "Merlin", "Northern Goshawk", "Northern Harrier", "Osprey", "Peregrine Falcon", "Prairie Falcon", "Red-shouldered Hawk", "Red-tailed Hawk", "Rough-legged Hawk", "Sharp-shinned Hawk", "Swainson's Hawk", "Turkey Vulture", "White-tailed Kite")) |> 
  # change name of goshawk
  mutate(CommonName = case_when(CommonName == "Northern Goshawk" ~ "American Goshawk",
                              TRUE ~ CommonName))


# Group by date and species to get a daily total count 
hw_full_dailycount <- hw_full_clean |> 
  group_by(ProtocolName, SiteName, SurveyYear, SamplingDate, CommonName) |> 
  summarise(DailyCount = sum(BirdCount, na.rm = TRUE), .groups = "drop")
```

### Banding data

```{r}
#| output: false
#| quiet: true
#| code-fold: true
#| code-summary: "Show the code"

# View dataset 
glimpse(banding_raw)

# View data elements 
#qc_summary(banding_raw)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# Clean banding data 
banding_clean <- banding_raw |> 
  # clean column names 
  clean_names(case = "upper_camel") |> 
  # select relevant columns 
  select(c(ProjectName, ProtocolName, SamplingDate, Site, BandingCode, BandingStatus, Species, Age, Sex, Morph)) |> 
  # rename Species to SpeciesCode
  rename("SpeciesCode" = "Species") |> 
  # create new column CommonName
  mutate(CommonName = case_when(SpeciesCode == "COHA" ~ "Cooper's Hawk",
                                SpeciesCode == "RTHA" ~ "Red-Tailed Hawk",
                                SpeciesCode == "AMKE" ~ "American Kestrel",
                                SpeciesCode == "SSHA" ~ "Sharp-shinned Hawk",
                                SpeciesCode == "NOHA" ~ "Northern Harrier",
                                SpeciesCode == "NOGO" ~ "Northern Goshawk",
                                SpeciesCode == "GHOW" ~ "Great Horned Owl",
                                SpeciesCode == "BWHA" ~ "Broad-winged Hawk",
                                SpeciesCode == "RSHA" ~ "Red-shouldered Hawk",
                                SpeciesCode == "PEFA" ~ "Peregrine Falcon",
                                SpeciesCode == "GOEA" ~ "Golden Eagle",
                                SpeciesCode == "FEHA" ~ "Ferruginous Hawk",
                                SpeciesCode == "RLHA" ~ "Rough-legged Hawk",
                                SpeciesCode == "PRFA" ~ "Prairie Falcon",
                                SpeciesCode == "MERL" ~ "Merlin",
                                SpeciesCode == "SWHA" ~ "Swainson’s Hawk",
                                SpeciesCode == "LEOW" ~ "Long-eared Owl",
                                SpeciesCode == "EUKE" ~ "Eurasian Kestrel",
                                SpeciesCode == "TUVU" ~ "Turkey Vulture",
                                SpeciesCode == "UNID_ACCP" ~ "Unidentified Accipiter",
                                SpeciesCode == "UNID_BUTEO" ~ "Unidentified Buteo",
                                SpeciesCode == "UNID_RAP" ~ "Unidentified Raptor",
                                SpeciesCode == "UNID_FALC" ~ "Unidentified Falcon",
                                SpeciesCode == "OSPR" ~ "Osprey",
                                SpeciesCode == "UNID_EAGLE" ~ "Unidentified Eagle",
                                TRUE ~ NA_character_
                                )) |> 
  # rename values in Sex column 
  mutate(Sex = case_when(Sex == "F" ~ "Female", 
                         Sex == "M" ~ "Male", 
                         Sex == "U" ~ "Unknown", 
                         is.na(Sex) ~ "Unknown",
                         TRUE ~ "Unknown")) |> 
  # remove rows where species code = NA 
  filter(!is.na(SpeciesCode))


```

## Color palette for plots

```{r}
parks_pal <- c("Blue Green" = "#4d9cbc",
               "Cerulean" = "#326e85",
               "Honey Bronze" = "#e5b03f",
               "Goldenrod" = "#d79d1d",
               "Rosy Copper" = "#dd5e4c",
               "Burnt Tangerine" = "#ce3b27",
               "Ocean Mist" = "#6bbaa9",
               "Seagrass" = "#489987",
               "Lime Moss" = "#adb449",
               "Palm Leaf" = "#8b913b")

# Repeat your palette to the needed length
parks_pal_expanded <- rep(unname(parks_pal), length.out = 20)

monochromeR::view_palette(parks_pal)


```

## Hawkwatch species counts and diversity

### Lollipop chart to show counts from 2020 to 2025

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Clean data 
hw_lolli <- hw_full_dailycount |> 
  # filter to years 2020 through 2025 
  filter(SurveyYear >= 2000) |> 
  # Group by common name 
  group_by(CommonName) |>
  summarise(Total = sum(DailyCount, na.rm = TRUE), .groups = "drop") |> 
  arrange(-desc(Total)) |> 
  # factor order by name to maintain descending order of counts 
  mutate(CommonName = factor(CommonName, levels=CommonName))
```

```{r, fig.width=8, fig.height=6}
#| code-fold: true
#| code-summary: "Show the code"
# Initalize ggplot 
lollipop <- ggplot(hw_lolli,  aes(x = CommonName, y = Total)) +
  # geoms for lollipop chart 
  geom_segment(aes(xend = CommonName, yend = 0), linewidth = 1.6,  color= parks_pal["Rosy Copper"]) +
  geom_point(size = 3.5, color=parks_pal["Burnt Tangerine"]) + 
  
  # count labels to right of points
  geom_text(aes(label = comma(Total)), 
            hjust = 0,  # left-align so label grows to the right
            nudge_y = 0.02 * max(hw_lolli$Total, na.rm = TRUE),  # push right
            size = 7,
            family = "Lato",
            fontface = "bold",
            color = parks_pal["Burnt Tangerine"]) +
  
  # flip x y coords 
  coord_flip(clip = "off") + 
  # axis labels 
  labs(y = "Total Birds Counted",
      # title = "Which Raptors are Our Volunteers Seeing?",
      # subtitle = "Hawk Watch Counts from 2000 - 2025"
      ) + 
  scale_y_continuous(breaks = seq(0, max(hw_lolli$Total), by = 50000), 
                     labels = scales::comma,
                     # add a bit of room on the right 
                     expand = expansion(mult = c(0, 0.08))) +
  # theme elements 
  theme_minimal() + 
  theme(
    # remove y axis title & legend 
    axis.title.y = element_blank(),
    legend.position = "none",
   
    #plot.title = element_text(size = 20, family = "Lato", face = "bold"),
    #plot.subtitle = element_text(size = 15, family = "Lato"),
    axis.title.x = element_text(family = "Lato", size = 20, margin = margin(t = 10, r = 0, b = 0, l = 0)),
    axis.text = element_text(family = "Lato", size = 20, color = "black"), 
    
    # remove background elements 
    panel.background = element_rect(fill = NA, color = NA),
    plot.background  = element_rect(fill = NA, color = NA),
    legend.background = element_rect(fill = NA, color = NA)
    )

lollipop
```


```{r}
# save as png
ggsave(
  filename = here::here("plot_exports", "hawkwatch_lollipop_plot.png"),
  plot = lollipop, width = 8, height = 6, units = "in", dpi = 300
)
# save as svg file
ggsave(
  filename = here::here("plot_exports", "hawkwatch_lollipop_plot.svg"),
  plot = lollipop,
  device = svg,
  width = 16.8,
  height = 8
)
```

Plot title: Which Raptors Are Our Volunteers Seeing? Subtitle: Hawkwatch counts of birds of interest from 2000 to 2025

## Hawkwatch Arrivial of Species

### Density plots to compare arrivial of Sharp-shinned hawks

This example looks at the density of sightings of some migratory species

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# Clean daily counts data to prepare for density plot 

# determine year range of ssh 
hw_density <- hw_full_dailycount |> 
  # filter for species of interest 
  filter(CommonName %in% c("Sharp-shinned Hawk", "Red-tailed Hawk", "Merlin", "Broad-winged Hawk")) |> 
  # filter for 1995 to 2024 
  filter(SurveyYear %in% 1995:2024) |> 
  # extract month and day from date 
  mutate(MonthDay = make_date(2000, month(SamplingDate), day(SamplingDate))) |> 
  # create year groups 
  mutate(YearGroup = case_when(SurveyYear %in% 1995:1999 ~ "1995-1999",
                               SurveyYear %in% 2000:2004 ~ "2000-2004",
                               SurveyYear %in% 2005:2009 ~ "2005-2009",
                               SurveyYear %in% 2010:2014 ~ "2010-2014",
                               SurveyYear %in% 2015:2019 ~ "2015-2019",
                               SurveyYear %in% 2020:2024 ~ "2020-2024")) |>
  # factor year group for plot order of y axis 
  mutate(YearGroup = fct_rev(YearGroup)) |> 

  # put the group you want on TOP as the LAST factor level
  mutate(CommonName = fct_relevel(CommonName,
                                 
                                  "Broad-winged Hawk",
                                  "Sharp-shinned Hawk", 
                                  "Merlin", 
                                  "Red-tailed Hawk", 
                                  ))


```

```{r, fig.width=8, fig.height=8}
#| code-fold: true
#| code-summary: "Show the code"

# Create density plot of bird observations
density_plot <- ggplot(
  hw_density,
  aes(
    x = MonthDay,
    y = YearGroup,
    fill = CommonName,  # fill color by species
    color = CommonName, # outline color by common species 
    weight = DailyCount # include weight for counts
  )  
) +
  # initialize density ridge
  geom_density_ridges(
    position = "identity", # order of species 
    alpha = 0.6, # transparency
    scale = 0.9, # vertical overlap (smaller value = less overlap)
    size = 0.6,               # thickness of the outline
    outline_type = "full"     # draw a full closed outline

  ) +
  # add custom colors per species 
  scale_fill_manual(values = c(
    "Sharp-shinned Hawk" = "#4F83DA",
    "Red-tailed Hawk"    = "#dd5e4c",
    "Broad-winged Hawk" = "#6bbaa9",
    "Merlin" = "#adb449"
    
  )) +
  
  scale_color_manual(values = c(
    "Sharp-shinned Hawk" = "#4F83DA",
    "Red-tailed Hawk"    = "#dd5e4c",
    "Broad-winged Hawk"  = "#6bbaa9",
    "Merlin"             = "#adb449"
  )) +
  
  # labels 
 # labs(title = "Density of Sightings During Hawk Watch Season",
       #subtitle = "Something about red-tailed seen consistantly but sharp-shinned \nseen more consistantly in September but in more recent years the peak is in October"
 #      ) +
  
  # set custom time range
  scale_x_date(
      limits = c(as.Date("2000-07-01"), as.Date("2000-12-30")),
      date_breaks = "1 month",
      date_labels = "%B"
    ) +

  # theme elements
  theme_minimal() + 
  theme(
    legend.position = "top", 
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.text = element_text(family = "Lato", size = 20),
    
    # remove axis titles 
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    
    # plot text elements
 #   plot.title = element_text(size = 20, family = "Lato", face = "bold"),
 #   plot.subtitle = element_text(size = 15, family = "Lato"),
    axis.text = element_text(family = "Lato", size = 20, face = "bold"),
 
    # remove background elements 
    panel.background = element_rect(fill = NA, color = NA),
    plot.background  = element_rect(fill = NA, color = NA),
    legend.background = element_rect(fill = NA, color = NA)

    )

density_plot
```


```{r}
# save as png
 ggsave(
  filename = here::here("plot_exports", "hawkwatch_density_plot.png"),
  plot = density_plot, width = 8, height = 8, units = "in", dpi = 300
)
# save as svg file
ggsave(
  filename = here::here("plot_exports", "hawkwatch_density_plot.svg"),
  plot = density_plot,
  device = svg,
  width = 14,
  height = 14
)
```

Plot title: plot subtitle: Each ridge shows the distribution of observation dates of each bird species. A a taller peak means the dates are more tightly concentrated around that time.

Context for this density plot:

This kind of plot reflects distribution of observation dates, not the actual count of birds.

"Each ridge shows the estimated distribution of observation dates for a species in that year group, normalized to area = 1 for that species–year group (weights influence shape, but the total area of the curve is still 1). The result is a smooth curve that answers “When during the season do sightings tend to cluster?”

Because density curves are normalized to integrate to 1, a taller peak means the dates are more tightly concentrated around that time. It doesn’t mean there were more total birds—just that the sightings (weighted by counts) are more focused in time.

So when looking at heights of ridges: Peaks show timing and concentration (weighted by Daily Count); heights are normalized per species–year group and do not reflect totals.

This plot does not answer which bird was seen the most often and instead addresses when during the season do sightings tend to cluster.

## Banding Data Plots

### Waffle chart

```{r}
#| code-fold: true
#| code-summary: "Show the code"
banding_waffle <- banding_clean |> 
  filter(BandingCode == 1) |> 
  # filter years (for plot test) 
  filter(SamplingDate >= ymd("2015-01-01")) |> 
  group_by(CommonName, Sex) |> 
  summarise(TotalBanded = n(), .groups = "drop") |> 
  # filter to big 4 species 
  filter(CommonName %in% c("Sharp-shinned Hawk", "American Kestrel", "Red-Tailed Hawk", "Cooper's Hawk")) |> 
  # order by smallest to largest 
  mutate(CommonName = forcats::fct_relevel(CommonName, c("American Kestrel", "Red-Tailed Hawk", "Sharp-shinned Hawk", "Cooper's Hawk")))
```

```{r, fig.width=8, fig.height=8}
#| code-fold: true
#| code-summary: "Show the code"
p_waffle <- ggplot(banding_waffle, aes(fill = Sex, values = TotalBanded)) +
  
  geom_waffle(color = "white", 
              size = 0.2,    # size of each square 
              n_rows = 20   # n_rows controls each panel's grid height
              ) +  
  # facet by species 
  facet_wrap(~ CommonName, 
             ncol = 1, 
             strip.position = "top", # position of species name 
             ) + 
  
  coord_equal() +                     # ensures squares are square
  scale_x_discrete() +                # avoids weird x scale expansion
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) + # increase space between plot and title
  
  # set custom colors 
  scale_fill_manual(values = c(
    "Female" = "#326e85",
    "Male" = "#adb449",
    "Unknown" = "grey50"
  )) +
  
  # labels 
#  labs(title = "Distribution of Select Banded Raptors From 2015-2025") + 
  
  # legend guide 
  guides(fill = guide_legend(title = "Sex")) +
 
   # theme elements 
  theme_void() +
  theme(
    # legend 
    legend.title = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(family = "Lato", size = 20),
    
    # strip text 
    strip.placement = "outside",
    strip.background.x = ggh4x::element_part_rect(
        fill = NA,
        colour = "grey30",   # color of the separator line
        linewidth = 0.6,
        side = "b",          # draw only the bottom border
   
      ),

    strip.text = element_text(face = "bold", family = "Lato",
                              margin = margin(t = 4, r = 6, b = 8, l = 6),
                              size = 20), 
    
 #   plot.title = element_text(size = 20, family = "Lato", face = "bold"),
    # axis text 
    axis.text = element_text(family = "Lato", size = 14), 
    
    # remove background elements 
    panel.background = element_rect(fill = NA, color = NA),
    plot.background  = element_rect(fill = NA, color = NA),
    legend.background = element_rect(fill = NA, color = NA)
  )

p_waffle


```

```{r}
# save as png
ggsave(
  filename = here::here("plot_exports", "banding_waffle_plot.png"),
  plot = p_waffle, width = 8, height = 8, units = "in", dpi = 300
)
# save as svg file
ggsave(
  filename = here::here("plot_exports", "banding_waffle_plot.svg"),
  plot = p_waffle,
  device = svg,
  width = 12,
  height = 16
)
```

Plot title: Who We Banded: Sex Distribution Across Four Raptor Species Plot subtitle: 2015 - 2025 

### Donut charts

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Clean data for donut charts 
banding_donut <- banding_clean |>
  # filter to new bands 
  filter(BandingCode == 1) |> 
  # group by common name 
  group_by(CommonName) |> 
  summarise(TotalCount = n(), .groups = "drop") |> 
  # Lump species outside top 6 (by weighted counts) to "Other Birds"
  mutate(CommonName = fct_lump_n(CommonName, n = 6,
                                 w = TotalCount,
                                 other_level = "Other Birds")) |> 
  # group by CommonName to group other birds 
  group_by(CommonName) |> 
  summarise(TotalCount = sum(TotalCount))

# Create dataset for Other Birds 
top_6_birds <- banding_donut |> 
  pull(CommonName)
top_6_birds <- as.character(top_6_birds)

banding_other_birds <- banding_clean |> 
  # filter to banding code = 1 
  filter(BandingCode == 1) |> 
  # group by common name 
  group_by(CommonName) |> 
  summarise(TotalCount = n(), .groups = "drop") |> 
  # filter to species not in top 6 
  filter(!CommonName %in% top_6_birds)

```

```{r, fig.width=8, fig.height=6}
#| code-fold: true
#| code-summary: "Show the code"

# --- Data prep ---

# Define custom colors 
banding_pal <- c(
               "Cooper's Hawk" = "#8b913b",
               "Merlin" = "#4d9cbc",
               "Red-Tailed Hawk" = "#dd5e4c",
               "American Kestrel" = "#6bbaa9",
               "Red-shouldered Hawk" = "#adb449",
               "Sharp-shinned Hawk" = "#d79d1d",
               "Other Birds" = "#326e85")

# Define order of species in donut chart (first element will start at 12 o'clock by default)
banding_order<- c(
  "Red-Tailed Hawk", 
  "Other Birds",
  "Sharp-shinned Hawk",
  "Red-shouldered Hawk",
  "American Kestrel",
  "Cooper's Hawk", 
  "Merlin"
)

# Reorder rows to match desired sequence
banding_donut <- banding_donut |> 
  mutate(CommonName = forcats::fct_relevel(CommonName, banding_order)) |> 
  arrange(CommonName)


# Compute percentages
banding_donut$fraction <- banding_donut$TotalCount / sum(banding_donut$TotalCount)
# Compute the cumulative percentages (top of each rectangle)
banding_donut$ymax <- cumsum(banding_donut$fraction)
# Compute the bottom of each rectangle
banding_donut$ymin <- c(0, head(banding_donut$ymax, -1))
# Compute label position for species
banding_donut$labelPosition <- (banding_donut$ymax + banding_donut$ymin) / 2

banding_donut$count_label <- paste0(comma(banding_donut$TotalCount))
banding_donut$species_label <- banding_donut$CommonName

# Controls for donut thickness & label placement
inner_r <- 2.65
outer_r <- 3.6
label_r <- 3.6   # where labeling segment starts 
x_max   <- 4.8    # extend plot range so labels aren't clipped

# Angle in radians for each mid-slice (0 at 12 o'clock, increasing clockwise)
banding_donut$angle <- 2 * pi * banding_donut$labelPosition
banding_donut$hjust_side <- ifelse(sin(banding_donut$angle) > 0, 0, 1)

# Total value of banding counts for center text
total_val <- scales::comma(sum(banding_donut$TotalCount))


# ---- Initialize ggplot ---- 
donut_plot <- ggplot(banding_donut, aes(ymax = ymax, ymin = ymin, xmax = outer_r, xmin = inner_r, fill = CommonName)) +
  geom_rect() +
  coord_polar(theta = "y", clip = "off") +
  
  # Make room so the repelled labels can live outside the ring
  xlim(c(1, x_max)) +
  
  # species text outside of ring 
  ggrepel::geom_text_repel(
    data = banding_donut,
    aes(x = label_r, y = labelPosition, label = species_label, color = CommonName, hjust = hjust_side),
    # text elements 
    fontface = "bold", 
    family = "Lato", 
    size = 8,
    
    inherit.aes = FALSE, # dont inherit aesthetics from the global ggplot
    direction = "y", # restricts repelling to the vertical direction
    nudge_x = 0.35, # pushed labels outward (expands segment)
    segment.size = 0.7, # segment thickness 
    segment.color = "grey50", # segment color 
    min.segment.length = 0, # always draw lines (positive value = short lines omitted)
    box.padding = 0.25, # how much space around each label box 
    point.padding = 0.1, # padding around anchor point
    force = 1.0, # higher values push labels further apart 
    max.overlaps = Inf, # dont drop labels due to overcrowding
    seed = 123 # reproducible label placement 
    ) +

  # count labels within ring
  geom_text(
    x = (inner_r + outer_r) / 2, 
    aes(y = labelPosition, label = count_label),
    size = 5.5, color = "white", fontface = "bold", family = "Lato"
  ) +
  
  # center text 
  annotate(
    geom = "richtext",
    x = 1, y = 0,
    label = paste0(
       "<span style='font-size:32pt; color:#4768C2;'>", total_val, "<br>",
       "<span style='font-family:Lato; font-weight:700; font-size:26pt;color:#1F305E;'>",
      "Total Birds Banded<br>",
      "From 1983-2025<br>",
      "</span>"
       
    ),

    # remove the default label box
    fill = NA, label.color = NA, label.padding = grid::unit(c(0, 0, 0, 0), "pt"),
    fontface = "bold",
    hjust = 0.5, 
    #vjust = 0.5,
    lineheight = 1.2  # tighten (default is ~1.2)
  ) +


  # custom colors 
  scale_fill_manual(values = banding_pal) +
  scale_color_manual(values = banding_pal) + 
  
  # theme elements 
  theme_void() +
  theme(
    # extra room left/right so labels aren't cut off
    plot.margin = margin(10, 40, 10, 40),  
    # remove legend 
    legend.position = "none", 
    
    # remove background elements 
    panel.background = element_rect(fill = NA, color = NA),
    plot.background  = element_rect(fill = NA, color = NA),
    legend.background = element_rect(fill = NA, color = NA)
  ) 
donut_plot
```

```{r}
# save as png
ggsave(
  filename = here::here("plot_exports", "banding_donut_plot.png"),
  plot = donut_plot, width = 8, height = 6, units = "in", dpi = 300
)

# save as svg file
ggsave(
  filename = here::here("plot_exports", "banding_donut_plot.svg"),
  plot = donut_plot,
  device = svg,
  width = 16,
  height = 12
)
```

#### Other Birds donut

```{r, fig.width=8, fig.height=6}
#| code-fold: true
#| code-summary: "Show the code"

## INSERT CUSTOM COLORS / ORDER 

# Define order of species in donut chart (first element will start at 12 o'clock by default)
banding_other_order<- c(
  "Peregrine Falcon", 
  "Eurasian Kestrel",
  "Prairie Falcon", 
  "Ferruginous Hawk",
  "Turkey Vulture",
  "Golden Eagle", 
  "Broad-winged Hawk", 
  "Northern Goshawk", 
  "Osprey", 
  "Long-eared Owl", 
  "Rough-legged Hawk", 
  "Northern Harrier",
  "Great Horned Owl", 
  "Swainson’s Hawk"
  
)

# Reorder rows to match desired sequence
banding_other_birds <- banding_other_birds |> 
  mutate(CommonName = forcats::fct_relevel(CommonName, banding_other_order)) |> 
  arrange(CommonName)

# color palette 
other_birds_pal <- c("#4E9BBC","#0353a4","#023e7d","#002855","#001845","#001233","#33415c","#4E9BBC","#0353a4","#023e7d","#002855","#001845","#001233","#33415c")

# Compute percentages
banding_other_birds$fraction <- banding_other_birds$TotalCount / sum(banding_other_birds$TotalCount)
# Compute the cumulative percentages (top of each rectangle)
banding_other_birds$ymax <- cumsum(banding_other_birds$fraction)
# Compute the bottom of each rectangle
banding_other_birds$ymin <- c(0, head(banding_other_birds$ymax, -1))
# Compute label position for species
banding_other_birds$labelPosition <- (banding_other_birds$ymax + banding_other_birds$ymin) / 2

banding_other_birds$count_label <- paste0(comma(banding_other_birds$TotalCount))
banding_other_birds$species_label <- banding_other_birds$CommonName

# Controls for donut thickness & label placement
inner_r <- 2.4
outer_r <- 3.5
label_r <- 3.5  # where labeling segment starts 
x_max   <- 4.8    # extend plot range so labels aren't clipped

# Angle in radians for each mid-slice (0 at 12 o'clock, increasing clockwise)
banding_other_birds$angle <- 2 * pi * banding_other_birds$labelPosition
banding_other_birds$hjust_side <- ifelse(sin(banding_other_birds$angle) > 0, 0, 1)

# Total value of banding counts for center text
#total_val <- scales::comma(sum(banding_donut$TotalCount))


# ---- Initialize ggplot ---- 
donut_plot2 <- ggplot(banding_other_birds, aes(ymax = ymax, ymin = ymin, xmax = outer_r, xmin = inner_r, fill = CommonName)) +
  geom_rect() +
  coord_polar(theta = "y", clip = "off") +
  
  # Make room so the repelled labels can live outside the ring
  xlim(c(1, x_max)) +
  
  # species text outside of ring 
  ggrepel::geom_text_repel(
    data = banding_other_birds,
    aes(x = label_r, y = labelPosition, label = species_label, color = CommonName, hjust = hjust_side),
    # text elements 
    fontface = "bold", 
    family = "Lato", 
    size = 8,
    
    inherit.aes = FALSE, # dont inherit aesthetics from the global ggplot
    direction = "y", # restricts repelling to the vertical direction
    nudge_x = 0.35, # pushed labels outward (expands segment)
    segment.size = 0.7, # segment thickness 
    segment.color = "grey50", # segment color 
    min.segment.length = 0, # always draw lines (positive value = short lines omitted)
    box.padding = 0.25, # how much space around each label box 
    point.padding = 0.1, # padding around anchor point
    force = 1.0, # higher values push labels further apart 
    max.overlaps = Inf, # dont drop labels due to overcrowding
    seed = 123 # reproducible label placement 
    ) +

  # count labels within ring
  geom_text(
    x = (inner_r + outer_r) / 2, 
    aes(y = labelPosition, label = count_label),
    size = 4.5, color = "white", 
    fontface = "bold", 
    family = "Lato"
  ) +
  
  # center text 
  annotate(
    geom = "richtext",
    x = 1, y = 0,
    label = paste0(
      "<span style='font-family:Lato; font-weight:700; font-size:28pt; color:#326e85;'>",
      "644<br>",
      "Other Birds<br>",
      "Banded<br>",
      "</span>"
    ),

    # remove the default label box
    fill = NA, label.color = NA, label.padding = grid::unit(c(0, 0, 0, 0), "pt"),
    fontface = "bold",
    hjust = 0.5, 
    #vjust = 0.5,
    lineheight = 1.0  # tighten (default is ~1.2)
  ) +


  # custom colors 
  scale_fill_manual(values = other_birds_pal) +
  scale_color_manual(values = other_birds_pal) + 
  
  # theme elements 
  theme_void() +
  theme(
    # extra room left/right so labels aren't cut off
    plot.margin = margin(10, 40, 10, 40),  
    # remove legend 
    legend.position = "none", 
    
    # remove background elements 
    panel.background = element_rect(fill = NA, color = NA),
    plot.background  = element_rect(fill = NA, color = NA),
    legend.background = element_rect(fill = NA, color = NA)
  ) 

donut_plot2
```

```{r}
# save as png
ggsave(
  filename = here::here("plot_exports", "banding_donut_plot2.png"),
  plot = donut_plot2, width = 8, height = 6, units = "in", dpi = 300
)

# save as svg file
ggsave(
  filename = here::here("plot_exports", "banding_donut_plot2.svg"),
  plot = donut_plot2,
  device = svg,
  width = 16,
  height = 12
)
```

Plot title: Which Kinds of Birds Have We Banded? Plot subtitle: Count of banded raptors from 1983 to 2025

## Plot Archive (not included in poster)

## Heatmap to show distribution of counts

This is code found [in this blog post](https://restateinsight.com/posts/general-posts/2024-12-github-contributions-plot/) which outlines how to make a make a `geom_rtile` function.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# `%||%` <- function(a, b) {
#   if(is.null(a)) b else a
# }
# 
# GeomRtile <- ggplot2::ggproto(
#   "GeomRtile", 
#   statebins:::GeomRrect, # 1) only change compared to ggplot2:::GeomTile
#                      
#   extra_params = c("na.rm"),
#   setup_data = function(data, params) {
#     data$width <- data$width %||% params$width %||% resolution(data$x, FALSE)
#     data$height <- data$height %||% params$height %||% resolution(data$y, FALSE)
# 
#     transform(data,
#       xmin = x - width / 2,  xmax = x + width / 2,  width = NULL,
#       ymin = y - height / 2, ymax = y + height / 2, height = NULL
#     )
#   },
#   default_aes = ggplot2::aes(
#     fill = "grey20", colour = NA, size = 0.1, linetype = 1,
#     alpha = NA, width = NA, height = NA
#   ),
#   required_aes = c("x", "y"),
# 
#   # These aes columns are created by setup_data(). They need to be listed here so
#   # that GeomRect$handle_na() properly removes any bars that fall outside the defined
#   # limits, not just those for which x and y are outside the limits
#   non_missing_aes = c("xmin", "xmax", "ymin", "ymax"),
#   draw_key = draw_key_polygon
# )
# 
# geom_rtile <- function(mapping = NULL, data = NULL,
#                        stat = "identity", position = "identity",
#                        radius = grid::unit(6, "pt"), # 2) add radius argument
#                        ...,
#                        na.rm = FALSE,
#                        show.legend = NA,
#                        inherit.aes = TRUE) {
#   ggplot2::layer(
#     data = data,
#     mapping = mapping,
#     stat = stat,
#     geom = GeomRtile,
#     position = position,
#     show.legend = show.legend,
#     inherit.aes = inherit.aes,
#     params = rlang::list2(
#       radius = radius,
#       na.rm = na.rm,
#       ...
#     )
#   )
# }
```

```{r}
# #| code-fold: true
# #| code-summary: "Show the code"
# # Add date information
# hw_heatmap <- hw_full_dailycount |> 
#   mutate(
#     # Gets the number of the week in the year
#     n_week = week(SamplingDate),
#     # Gets weekday number - Starts the week at sunday
#     n_day = wday(SamplingDate, week_start = 7),
#     # Weekday labels for the plot
#     weekday_label = wday(SamplingDate, week_start = 7, label = TRUE, abbr = TRUE),
#     weekday_label = fct_rev(weekday_label),
#     # Month labels for the plot
#     month = month(SamplingDate, label = TRUE, abbr = TRUE) 
#     )
# 
# # Filter to year 2019 and Sharp-shinned Hawk for test 
# hw_heatmap_2019 <- hw_heatmap |> 
#   filter(SurveyYear == 2019, 
#          CommonName == "Sharp-shinned Hawk")
```

For 2019 only

```{r}
# #| code-fold: true
# #| code-summary: "Show the code"
# # Find the positions of the month labels
# tab <- hw_heatmap_2019|> 
#   summarise(nmin = min(n_week), .by = "month")
# 
# # heat map
# ggplot(hw_heatmap_2019, aes(n_week, weekday_label)) +
#   geom_rtile(
#     aes(fill = DailyCount),
#     color = "white",
#     radius = unit(2, "pt"),
#     width = 0.9,
#     height = 0.9
#     ) +
#   # Highlight the months on the horizontal axis
#   scale_x_continuous(
#     limits = c(27, 53), # from early july to late december 
#     breaks = tab$nmin,
#     labels = as.character(tab$month),
#     position = "top",
#     expand = c(0, 0)
#   ) +
#   # Highlight days of the week on the vertical axis
#   scale_y_discrete(breaks = c("Mon", "Wed", "Fri")) +
#   # Adjust color palette
#   scale_fill_distiller(
#     palette = "Oranges",
#     direction = 1,
#     na.value = "gray90") +
#   
#   # Removes x and y labels
#   labs(x = NULL, 
#        y = NULL,
#        title = "Sharp-shinned Hawk Counts in 2019") +
#   # Removes the color legend
#   guides(fill = "none") +
#   theme_minimal() +
#   theme(
#     panel.grid = element_blank(),
#     text = element_text(color = "gray10")
#   )
```

Facet wrap plot from to 2019 to 2024

```{r}
# #| code-fold: true
# #| code-summary: "Show the code"
# # Filter to 2019-2025 and Sharp-shinned Hawk for test 
# hw_heatmap_2019_2024 <- hw_heatmap |> 
#   filter(SurveyYear >= 2019 & SurveyYear <= 2024,
#          CommonName == "Sharp-shinned Hawk")  #"American Kestrel"  "Sharp-shinned Hawk"
```

```{r}
# #| code-fold: true
# #| code-summary: "Show the code"
# # Find the positions of the month labels (Not the best way to fix this, i just want to see )
# # tab <- hw_heatmap_2019_2024 |> 
# #   filter(SurveyYear == 2023) |> 
# #   summarise(nmin = min(n_week), .by = "month")
# 
# 
# min_year <- hw_heatmap_2019_2024 |>
#   arrange(n_week, SurveyYear) |>
#   slice(1) |>
#   pull(SurveyYear)
# 
# tab <- hw_heatmap_2019_2024 |>
#   filter(SurveyYear == min_year) |>
#   summarise(nmin = min(n_week), .by = month) |>
#   arrange(month)
# 
# 
# 
# # heat map
# plot <- ggplot(hw_heatmap_2019_2024, aes(n_week, weekday_label)) +
#   geom_rtile(
#     aes(fill = DailyCount),
#     color = "white",
#     radius = unit(2, "pt"),
#     width = 0.9,
#     height = 0.9
#     ) +
#   # Highlight the months on the horizontal axis
#   scale_x_continuous(
# #    limits = c(27, 53), # from early july to late december 
#     breaks = tab$nmin,
#     labels = as.character(tab$month),
#     position = "top",
#     expand = c(0, 0)
#   ) +
#   # Highlight days of the week on the vertical axis
#   scale_y_discrete(breaks = c("Mon", "Wed", "Fri")) +
#   # Adjust color palette
#   scale_fill_distiller(
#     palette = "Greens",
#     direction = 1,
#     na.value = "gray90") +
#   
#   # Removes x and y labels
#   labs(x = NULL, y = NULL) +
#   # Removes the color legend
#   guides(fill = "none") +
#   # Separate by year 
#   facet_wrap(~ SurveyYear, ncol = 1) +
#   
#   theme_minimal() +
#   theme(
#     panel.grid = element_blank(),
#     text = element_text(color = "gray10")
#   )
# 
# plot
# 
#  Add month labels per facet as text at the top
#  plot +
#    geom_text(
#      data = tab,
#      aes(x = nmin, y = levels(hw_heatmap_2019_2024$weekday_label)[1],  # top row #label
#          label = as.character(month)),
#      inherit.aes = FALSE,
#      vjust = -0.6,        # push above tiles
#      size  = 3.2
#    )

```

### Stacked bar chart to show sex and counts

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# Stacked bar chart 

# ggplot(banding_clean, aes(x = CommonName, fill = Sex)) +
#   geom_bar(position = "stack", color = "grey20", linewidth = 0.2) +
#   scale_fill_brewer(palette = "Set2") +
#   labs(
#     x = "Common Name",
#     y = "Number of records",
#     fill = "Sex",
#     title = "Number of records per species by sex"
#   ) +
#   theme_minimal(base_size = 12) +
#   
#   theme(
#     axis.text.x = element_text(angle = 45, hjust = 1),
#     legend.position = "top",
#     legend.title = element_blank(),
#     plot.title = element_text(size = 20, family = "Lato", face = "bold"),
#     axis.text = element_text(family = "Lato", size = 12)
#     
#     )

```

### Tree Map to show species diversity from 2020 to 2025

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# # Clean daily counts to prepare for treemap plot
# hw_treemap <- hw_full_dailycount |> 
#   # remove Monarch Butterfly  
#   filter(!CommonName ==  "Monarch Butterfly") |> 
#   # filter to years 2020 through 2025 
#   filter(SurveyYear >= 2020) |> 
#   # group by species 
#   group_by(CommonName) |> 
#   summarise(Total = sum(DailyCount)) |> 
#   # order by number 
#   arrange(desc(Total)) |> 
#   slice_head(n = 25)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# # Count how many species (levels) you have
# n_species <- n_distinct(hw_treemap$CommonName)
# 
# # Repeat your palette to the needed length
# parks_pal_expanded <- rep(unname(parks_pal), length.out = n_species)
# 
# # Get bird stats for subtitle 
# total_birds_20_25 <- comma(sum(hw_treemap$Total))
# total_bird_species <- length(unique(hw_full_dailycount$CommonName))
# 
# 
# 
# # Build tree map using treemapify 
# ggplot(hw_treemap, aes(area = Total, fill = CommonName, label = CommonName)) +
#   geom_treemap() +
#   geom_treemap_text(colour = "white",
#                     place = "centre",
#                     size = 15, 
#                     grow = TRUE,
#                     family = "Lato") +
#   # title and subtitle text 
#   labs(title = "Which birds are seen the most during Hawk Watch?",
#        subtitle = paste0("From 2020 to 2025 a total of ", total_birds_20_25, " birds were identified by volunteers spanning ", total_bird_species, " different species. The most common \nbird sited was the Band-tailed Pigeon, followed by Turkey Vultures and Red Tailed Hawk\n")) +
#   
#   # fill treemap with custom color palette called 'parks_pal'
#   scale_fill_manual(values = unname(parks_pal_expanded)) +
#   
#   # theme elements
#   theme(
#     # remove legend
#     legend.position = "none",    
#     # title to Lato & center
#     plot.title = element_text(family = "Lato", face = "bold", 
#                               size = 25),
#     # subtitle to Lato
#     plot.subtitle = element_text(
#       family = "Lato", size = 18, margin = margin(t = 6))
#     )
```
